Ornaments as indicators of Social change before and after European contact at Kiwulan, Northeastern Taiwan

### 2019-07-08 analysis/paper/paper.Rmd
add image to figure folder and some more editing

Taiwan had been involved in regional exchange network of East Asia and Southeast Asia since Neolithic and became part of a global trading network during the early 17th century due to western European expansion that connected European and Asia. Compared to other places in Asia and Oceania, the impact of European colonialism on indigenous communities in East Asia appears to have been much less pronounced. Direct colonial rule was rare and limited, but the question of long-lasting indirect impacts on local indigenous communities remains unanswered. Indirect effects of colonialism is emphasized increasingly to discuss the colonial impact on the indigenous people in the periphery of colonial control [@TorrenceandClarke2000; @Trabert2017]. In many parts of the world, foreign trade goods introduced into local indigenous societies in a colonial or imperial contexts might cause significant transformations of indigenous economic, cultural, and socio-political systems [@Dietler1997; @Junker1993; @Mitchell2000; @Silliman2005]. Contrary to as passive receptors of imperial power, the agency of indigenous people in the colonial contexts has been broadly discussed to emphasize the ability of active adoptions of foreign effects, negotiation between colonized and colonizer, and resistance of indigenous people through daily cultural practices [@Dietler2015; @Given2004; @Rubertone2000; @Silliman2001; @Torrence2000; @TorrenceandClarke2000; @Voss2005]. Studying the consumption of foreign goods is one way to understand indigenous experience of the colonial encounter and explore the agency of indigenous people by examining how those goods were distributed and used when obtained [@Dietler2005; @Mullins2011; @Scaramelli2005].

Despite colonization of Taiwan by the Spanish and the Dutch from 1624 to 1662, little has been discovered in archaeology about the The reaction of indigenous people to the European colonization. Northeastern Taiwan, Yilan, is an especially unique example for exploring indirect effects due to its physically isolated location that made it not easily accessible to the Europeans and the colonial control was less compared to other parts of this region [@Andrade2007; @Kang2012]. Although the location is relatively peripheral, the sea provides a way through which indigenous people could interact with other groups from different regions, including other native indigenous groups, foreign traders such as the Han people of China, and Europeans. Large-scale and frequent trade activities brought a wide variety of trade objects into indigenous communities in Yilan through local regional exchange network. Among those trade goods, ornament such as glass and stone beads was one of the common foreign goods found at the settlement sites in this region [@Chen2007; @Li2014; @NMTH2005]. Body adornment is usually viewed as a signal in indigenous social life to represent and reinforce their status through the body as the scene to display [@Joyce2005; @Scaramelli2005]. For example, the consumption of stone beads in Southeast Asia during Iron Age were usually associated with increasing social stratification or socio-political complexity [@Bellina2014; @Carter2016; @Francis2002; @Theunissen2000].

In this paper, I explore the evidence from a site that spans the time from late Iron Age to historical period by addressing the question of whether colonial influence on indigenous populations can be detected in ornaments recovered from the archaeological record at Kiwulan (1400-1900 AD), a large Iron Age settlement in northeastern Taiwan [@Chen2007]. Prior to the contact with the Europeans in the early 17th century, northern Taiwan had been involved in long-distance networks of East Asia since 14th century, and later encountered a large wave of Han Chinese migration in the 19th century [@Chen2005]. As @Lape2003 points out based on case studies in island Southeast Asia, the study of multiple episodes of culture contact has potential to expand our understanding of the causes and effects of culture contact both locally and worldwide. Northeastern Taiwan provides a good example to explore the indirect impacts on indigenous societies by comparing archaeological evidence from different episodes of culture contact based on a long-term perspective. The variety and cultural context of trade ornaments will be reviewed, and their capacity to inform on how the distribution of ornaments reflect social changes will be examined. My hypothesis is that there was a greater diversity of ornaments types and materials at Kiwulan after European contact due to the large scale exchange network, and the consumption of ornaments present some spatial pattern that indicates increasing social inequality. I hypothesize that after Chinese contact there was a decline in the production, use and discard of ornaments at Kiwulan.

@@

Before European arrival, the indigenous people in northern Taiwan had already established their own exchange network that defined as "inter-insular trade" by @Chen2005 [p. 12] to refer to small-scale regional trade on irregular basis . Despite being on the periphery of major trade ports at northern Taiwan, Yilan was involved in this network through the periodic exchange activities between different indigenous groups. Yilan, also called "Kavalan" by local indigenous people since the prehistoric time, is a alluvial plain circumscribed by the Pacific and mountains on the sides. Rivers and seas provide the way for the interaction between local indigenous people and another native group called Basay in northern Taiwan and Han Chinese. Kavalan people offered rice, deer hides, and gold in exchange for beads, metal tools or ornaments, ironwares, porcelains, and textiles with outside traders [@Chen2005; @Hsieh2009; @LiandWu2006]. This local exchange network of northeastern Taiwan had global connections after the 16th century because of its location on shipping routes between China and Japan, and was intensified in the 17th century when the arrival of Europeans and their trade goods [@Chen2005; @Wang2007].

Historical records from Spanish and Dutch provide rich information about the historical and cultural background both for Europeans and indigenous people in the early 17th century. Spanish occupied and built forts in Northern Taiwan since 1624, from which Spanish missionaries often went to indigenous settlement and left behind numerous records from those places. The report of the Dominican Jacinto Esquivel in 1632 mentioned that the Taparri, an indigenous tribe from northern Taiwan, usually used cuentas (agate beads) as a materials to exchange with necessities they need with other indigenous people. This trend then gradually spread to other areas, and even the Spanish soldiers used agate as bargaining chips for gambling, since agate equaled money [@LiandWu2006, pp. 132-149].  European contact was focused on trading with indigenous communities in northeastern Taiwan. Trading activity of many types of goode among these indigenous communities increased with the arrival of European colonizers. Thus, we predict that this trading activity resulted in an increase in the amount and diversity of ornaments coming into indigenous communities in northeastern Taiwan.

@Mullins2011---<They examine the ways gift-giving worked in the Middle Orinoco region of Venezuela, where missionaries hoped to secure indigenous alliances by providing goods for which they perceived a native demand. One class of goods missionaries provided was alcohol, which was used commonly in feasts and ritual contexts, but interjecting externally produced alcohol unseated existing production and consumption systems. Prior to colonization, drink had been produced by indigenous domestic units to enhance a family's prestige, but the introduction of imported spirits compelled hosts to obtain them through exchange, which forced increasing reliance on the production of cash crops and increased debt among indigenous consumers. This shift rendered drink “a commodity—a product that had to be bought, rather than produced,” so even though European alcohol was consumed within existing consumer contexts it led to dependency and exploitation>

Father Jacinto Esquivel recorded how people in Tamsui entered marriages. A man who wanted to take a wife had to give her parents cuentas. If a married man was found having cheated on his wife, he only had to pay the witness with agate or golden beads and the matter was settled [@LiandWu2006, pp. 151]. Other records mention that the female shamans (majuorbol) in the tribe used agate beads for healing ceremonies in rituals [@Pao2008, pp. 122-151]. Regarding funerals, some indigenous people buried the dead under their houses or in the vicinity. They put a mat into the grave to protect the body from moisture and then put cooked rice on both sides of the head as provision. The deceased were buried in small graves with both knees bent, and on the grave they placed quivers with arrows, pottery, agate beads and other things that they would need. The greater the influence of the deceased, the more agate, pottery and cloths were placed on the grave [@LiandWu2006, pp. 153].

Documents from the Qing dynasty usually describe the decorative purposes of the ornaments. In the Kavalan Subprefecture Gazetteer (Chen2006[1852]: 308), the Kavalan Zhi Lue (Ko1993[1837]: 11, 126) and the Dong Cha Ji Lue (Yao1996[1829]: 77), it is mentioned that aborigines in Yilan used metal threads to weave so called golden carp-shaped ornaments. Their number was small and their value high, hence only the rich possessed them. Ordinary people wore agate beads or glass beads on their head or neck during various festivals. In 1895, at the beginning of Japanese rule over Taiwan, Ino Kanori [@Kanori1996, pp. 227-232] made field surveys among various ethnic groups in Taiwan. He describes the Plains Aborigines from Yilan as not using carp-shaped ornaments any more, but mentions that older people still use beads ornaments.

Recent ethnographic research of the Kavalan tribe describes agate beads as used in divination ceremonies, the subli, by female shamans [@Liu2008, pp. 133-134]. Interview records describe agate beads as valuable objects passed down from mother to daughter. When they were not used, they were often put on roof beams so that they were not seen by other people. But the origin of these beads is still not known. Later, because they were bought by Japanese and people in Taipei, agate beads became rarer [@Hsu1992, pp. 22] and carp-shaped ornaments had completely disappeared. Modern ethnographic research shows that agate beads were not only used as decoration, but also played an important role in divination. It was usually the female shamans that were responsible for divination rites, a technique passed down by ancestors. Most aborigines today don’t know the origin of agate beads any more, but their worth is still highlighted by oral history and through their scarcity.

Those historical documents from the early 17th century to the modern ethnography described how local indigenous people use those ornaments in local cultural contexts that represents some social roles or status. However, compared to European contact period, there is less mention of beads in Chinese contact period and the description about ornaments is only limited to their dressing culture.

@@

```{r, KWL-map, fig.cap = "Kiwulan location in Southeast Asia", out.width= "400px", echo=FALSE}
knitr::include_graphics("kiwulan-location-map.png")
```

Kiwulan \@ref(fig:KWL-map) is located at northern Yilan and was excavated from 2001 to 2004 that creates 262 squares with the total area of 3,814 meter squares. Most beads were found in situ while digging and some were collected through screens with 20 mm and 1.5 mm mesh. The archaeological evidence includes a rich amount of artifacts, burials, middens, post holes, wooden pillars, and stone structures. The chronology of Kiwulan can be divided into Lower Layer Culture (700 - 1200 AD) and Upper Layer Culture (1400 - 1900 AD) with a sterile layer in between based on a series of 32 radiocarbon dates, which coordinate to the late Iron Age and Proto-historical period in Taiwan. This paper focuses on the Upper Layer Culture which overlaps with the contact periods with the Europeans and Chinese.

@@

```{r read-in-the-data}
library(here)
library(tidyverse)
library(ggplot2)
kwl_upper <- readxl::read_excel(here("analysis", "data", "raw_data", "Kiwulan_Ornament_Upper.xlsx"))
```

@@

ornaments_shape <-
  ornaments_period_clean %>% 
  group_by(period, Categories, Type) %>% 
  tally() %>% 
  spread(period, n, fill = 0) %>% 
  filter(!Categories %in% c("Unknown Metal", 
                            "Perforated plate", 
                            "Knited item",
                            "Hair pin",
                            "Bone Ornament",
                            "Metal Piece",
                            "Glass ring"))

@@

Here we use Monte Carlo method to test for the complete spatial randomness of spatial events based on the average nearest-neighbor distance method. We hypothesize that the distribution of ornaments is consistent with a completely random process, and then simulate the presence of ornaments across the space for one thousand times for each time periods. We compare our data to the random pattern generated by simulated processes to see if the our data is randomly distributed or not. The results \@ref(fig:hypo-test-distrubution) are represented by histogram showing one thousand simulation with the blue line indicating our observed value. The results show that 100% of the simulated values are much more greater than our observed ANN value during the European contact period, which means the ornaments are clustered distributed. Similar result are also observed during the pre-European period but less extreme. The distribution of ornaments was random during the Chinese contact period and about one third of the simulated values are greater than our observed ANN value. This testing explains that the distribution of ornaments during the pre-European and European contact period might reflect the presence of different social groups. Moreover, the cluster of ornaments way more significant during the European contact period that might indicate some extent of control of ornaments.

>>>>

Taiwan had been involved in regional exchange network of East Asia and Southeast Asia since the tenth century and became part of a global trade systmes during the early 17th century due to western European expansion that connected European and Asia [@Liu2017]. Compared to other places in Asia and Oceania, the impact of European colonialism on indigenous communities in East Asia appears to have been much less pronounced. Direct colonial rule was rare and limited, but the question of long-lasting indirect impacts on local indigenous communities remains unanswered. Indirect effects of colonialism is emphasized increasingly to discuss the colonial impact on the indigenous people in the periphery of colonial control [@TorrenceandClarke2000; @Trabert2017]. In many parts of the world, foreign trade goods introduced into local indigenous societies in a colonial or imperial contexts might cause significant transformations of indigenous economic, cultural, and socio-political systems [@Dietler1997; @Junker1993; @Mitchell2000; @Silliman2005]. Contrary to as passive receptors of imperial power, the agency of indigenous people in the colonial contexts has been broadly discussed to emphasize the ability of active adoptions of foreign effects, negotiation between colonized and colonizer, and resistance of indigenous people through daily cultural practices [@Dietler2015; @Given2004; @Rubertone2000; @Silliman2001; @Torrence2000; @TorrenceandClarke2000; @Voss2005]. Studying the consumption of foreign goods is one way to understand indigenous experience of the colonial encounter and explore the agency of indigenous people by examining how those goods were distributed and used when obtained [@Dietler2005; @Mullins2011; @Scaramelli2005].

Despite colonization of Taiwan by the Spanish and Dutch from 1624 to 1662, little had been discussed in archaeology about the reaction of indigenous people to the European colonization. @Berrocal2018 study the long-term record of Heping Dao in northern Taiwan and explore the effects of European colonization on the local populations by focusing on the environmental change and introduction of diseases, with a little mention of indigenous social complexity. Northeastern Taiwan, Yilan, is an especially unique example for exploring indirect effects due to its physically isolated location that made it not easily accessible to the Europeans and the colonial control was less compared to other parts of this region [@Andrade2007; @Kang2012]. Although the location is relatively peripheral, the sea provides a way through which indigenous people could interact with other groups from different regions, including other native indigenous groups, foreign traders such as the Han people of China, and Europeans. Large-scale and frequent trade activities brought a wide variety of trade objects into indigenous communities in Yilan through local regional exchange network. Among those trade goods, ornament such as glass and stone beads was one of the common foreign goods found at the settlement sites in this region [@Chen2007; @Li2014; @NMTH2005]. Body adornment is usually viewed as a signal in indigenous social life to represent and reinforce their status through the body as the scene to display [@Joyce2005; @Scaramelli2005]. For example, the consumption of stone beads in Southeast Asia during Iron Age were usually associated with increasing social stratification or socio-political complexity [@Bellina2014; @Carter2016; @Francis2002; @Theunissen2000].

In this paper, I explore the evidence from a site spanning from late Iron Age to historical period by addressing the question of whether colonial influence on indigenous populations can be detected in ornaments recovered from the archaeological record at Kiwulan (1400-1900 AD), a large Iron Age settlement in northeastern Taiwan [@Chen2007]. Prior to the contact with the Europeans in the early 17th century, northern Taiwan had been involved in both regional and long-distance East and Southeast Asian trade systems, and later encountered a large wave of Han Chinese migration in the 19th century [@Chen2005]. As @Lape2003 points out based on case studies in island Southeast Asia, the study of multiple episodes of culture contact has potential to expand our understanding of the causes and effects of culture contact both locally and worldwide. Northeastern Taiwan provides a good example to explore the indirect impacts on indigenous societies by comparing archaeological evidence from different episodes of culture contact based on a long-term perspective. The variety and cultural context of trade ornaments will be reviewed, and their capacity to inform on how the distribution of ornaments reflect social changes will be examined. I hypothesize that there was a greater diversity of ornaments types and materials at Kiwulan after European contact due to the large scale exchange network, and the consumption of ornaments present some spatial pattern that indicates increasing social inequality. I hypothesize that after Chinese contact there was a decline in the production, use and discard of ornaments at Kiwulan.

@@

Prior to European colonization, the indigenous people in northern Taiwan had already established their own exchange network that defined as "inter-insular trade" by @Chen2005 [p. 12] to refer to small-scale regional trade between China and Taiwan on irregular basis. This local exchange network of northeastern Taiwan had increasingly influenced by long distance trade between Fuzhou or Quanzhou in China and Ryukyu in Japan since 15th century due to its location on the shipping routes, and was intensified in the 17th century when the arrival of the Europeans and their trade goods [@Chen2005; @Liu2017; @Wang2007]. Despite being on the periphery of major trade ports, Keelung and Tanshui, in northern Taiwan, Yilan was involved in this network through the periodic exchange activities between different indigenous groups. Yilan, also called "Kavalan" by local indigenous people since the prehistoric time, is a alluvial plain circumscribed by the Pacific and mountains on the sides. Rivers and seas provide the way for the interaction between local indigenous people and another native group called Basai in northern Taiwan and Han Chinese. Kavalan people offered rice, deer hides, and gold in exchange for beads, metal tools or ornaments, ironwares, porcelains, and textiles with outside traders [@Chen2005; @Hsieh2009; @LiandWu2006].

Historical documents from the Spanish and Dutch in the early 17th century provide some information about the social life and culture of indigenous people in Northern Taiwan when the Europeans arrived. The Spanish founded Fort San Salvador at Keelung in 1626 and Fort San Domingo in 1629 at Tamsui, and sent missionaries to local indigenous settlements in this region for religious purpose [@Blusse2000, pp. 343]. The report of the Dominican priest Fr. Jacinto Esquivel in 1632 mentioned that the Taparri, an indigenous tribe from northern Taiwan, usually used cuentas (agate beads) in exchange for necessities with other indigenous group. This form of exchange spread widely and even the Spanish soldiers learned to use agate as bargaining chips for gambling, since agate equaled money [@LiandWu2006, pp. 132-149]. Father Jacinto Esquivel also recorded how indigenous viewed it as prestige goods in their culture. A indigenous male who wanted to get married had to pay agate beads to the parents of his future wife. Also, agate or golden beads could be used as to solve conflicts in their daily lives [@LiandWu2006, pp. 151]. Other records mention that the female shamans (majuorbol) in the tribe would use agate beads as magical items in ritual practice for body healing[@Borao2009, pp. 122-151]. A funeral process in indigenous culture was recorded about how agate beads used commonly in ritual contexts, where more agate beads, pottery and cloths were be placed into the grave of more influential people to indicate a family's prestige [@LiandWu2006, pp. 153].

Those documents indicate that agate beads or golden beads were already treated as prestige goods before the arrival of the Europeans, that is believed to be introduced earlier from Chinese traders. In 1642, the Dutch VOC defeated the Spanish and take over the northern Taiwan. They introduced feudal symbolism in an attempt to control the indigenous communities by asking indigenous leaders to attend annual ceremony for feudal practices [@Andrade2007, ch. 9; @Kang2016, ch. 4]. More beads entered indigenous settlements through the gift-giving that the Dutch provided goods based on the demand of native people to secure indigenous alliances in the annual ceremony or on the way of traveling [@Kang2016, ch. 6]. Thus, I assume that the way of gift giving resulted in an increase in the amount and diversity of ornaments in northeastern indigenous communities within existing consumer contexts during this period.

The documents written by Chinese historians in Qing dynasty usually described the decorative purposes of the ornaments and some documents had similar texts that might be copied from previous ones. According to those records, [@Chen2006[1852]: 308], [@Ko1993[1837]: 11, 126] and [@Yao1996[1829]: 77], it is mentioned that indigenous people in Yilan used ornaments in ceremonial contexts to display their status. Amoung those ornaments, golden carp-shaped ornaments made from metal threads had higher value due to its delicacy and rare material that only possessed by rich people. Other people wore agate beads or glass beads on their head or neck to participating various ceremonies. In 1895, at the beginning of Japanese colonization, Ino Kanori [@Kanori1996, pp. 227-232] conducted a field survey for several ethnic groups and found that carp-shaped ornaments had not been used in Yilan that time, but elder people still used beads as ornaments.

Recent ethnographic research of the Kavalan tribe describes that agate beads were used in divination practice, called "subli", by female shamans [@Liu2008, pp. 133-134]. Interview records mention agate bead is valuable object that are usually passed down from mother to daughter as a heirloom in shaman families without knowing the exact origin. Were not been using, beads would be often hided at invisible place such as on the roof beams to make sure no one sees that. During the Japanese colonization period, the number of agate beads declines due to the trade with Japanese who live in Taipei for collecting purpose [@Hsu1992, pp. 22] and carp-shaped ornaments had completely disappeared. Modern ethnographic research shows that agate beads were not only used as decoration, but also played an important role in divination that is consistent with the observation in European colonial period. It was usually the female shamans conducting divination ritual that passed down from their ancestors. Despite most aborigines today don’t know the origin of agate beads, the value of beads is still be highlighted through oral history and their scarcity. Those historical documents from the early 17th century to the modern ethnography described how local indigenous people use ornaments in local cultural contexts that represents some social roles or status. However, compared to European contact period, there is less mention of beads in Chinese contact period and the description about ornaments is only limited to their dressing culture.

@@

```{r, KWL-map, fig.cap = "Kiwulan location in Southeast Asia", out.width= "400px", echo=FALSE}
library(here)
knitr::include_graphics(here("analysis", "figures", "kiwulan-location-map.png"))
```

excavate an open area for the first time reaching a surface a bit larger than 109 m2.

Kiwulan \@ref(fig:KWL-map) is located at northern Yilan and was excavated from 2001 to 2004 that creates an open area with 262 squares (4m by 4m) reaching 3,814 meter squares in total. Most beads were found in situ while digging and some were collected through screens with 20 mm and 1.5 mm mesh. The archaeological evidence includes a rich amount of artifacts, burials, middens, post holes, wooden pillars, and stone structures. The chronology of Kiwulan can be divided into Lower Layer Culture (700 - 1200 AD) and Upper Layer Culture (1400 - 1900 AD) with a sterile layer in between based on a series of 32 radiocarbon dates, which coordinate to the late Iron Age and Proto-historical period in Taiwan. This paper focuses on the Upper Layer Culture which overlaps with the contact periods with the Europeans and Chinese.

@@

```{r read-in-the-data}
library(tidyverse)
library(ggplot2)
kwl_upper <- readxl::read_excel(here("analysis", "data", "raw_data", "Kiwulan_Ornament_Upper.xlsx"))
```

@@

ornaments_shape <-
  ornaments_period_clean %>% 
  group_by(period, Categories, Type) %>% 
  tally() %>% 
  spread(period, n, fill = 0) %>% 
  filter(!Categories %in% c("Unknown Metal", 
                            "Perforated plate", 
                            "Knited item",
                            "Hair pin",
                            "Bone Ornament",
                            "Metal Piece",
                            "Glass ring")) %>%
  filter(!Type %in% c("other",
                      "unclassified")) %>% 
  arrange(Categories, desc(`European Contact`))

@@

Here we use Monte Carlo method to test for the complete spatial randomness of spatial events based on the average nearest-neighbor distance method. We hypothesize that the distribution of ornaments is consistent with a completely random process, and then simulate the presence of ornaments across the space for one thousand times for each time periods. The data was compared to the random pattern generated by simulated processes to see if the our data is randomly distributed or not. The results \@ref(fig:hypo-test-distrubution) are represented by histogram showing one thousand simulation with the blue line indicating our observed value. The results show that 100% of the simulated values are much more greater than our observed ANN value during the European contact period, which means the ornaments are clustered distributed. Similar result are also observed during the pre-European period but less extreme. The distribution of ornaments was random during the Chinese contact period and about one third of the simulated values are greater than our observed ANN value. This testing explains that the distribution of ornaments during the pre-European and European contact period might reflect the presence of different social groups. Moreover, the cluster of ornaments way more significant during the European contact period that might indicate some extent of control of ornaments.
